# Comparison of ORA and GSEA Workflows for RNA-seq Enrichment Analysis in R {-}

## Introduction {-}

During DSFB1 course, I have learned how to perform ORA Gene Enrichment analysis for an RNA-sequencing experiment and applied it on the provided dataset for the final assessment. 

To further deepen my knowledge of RNA analysis, I want to learn how to perform another type of RNA-seq enrichment analysis: GSEA. 

In this self-study assignment, I want to: 

- Learn how enrichment methodology affect the biological interpretation of RNA-seq data and compare the previous ORA results with a new GSEA analysis using the same DESeq2 dataset. 

- Obtain a more complete picture of the previously used data.

## Background | Difference between GSEA and ORA {-}

Difference between GSEA(Gene Set Enrichment Analysis) and ORA(Over-Representation Analysis): 
ORA uses a threshold-filtered list of significant genes, while GSEA uses all genes ranked by expression change. ORA is ideal for quick, high-magnitude gene lists, whereas GSEA excels at identifying subtle, coordinated changes in entire pathways.

https://pluto.bio/resources/Learning%20Series/gsea-vs-ora-two-key-pathway-analysis-approaches-for-next-gen-sequencing-data


## Data description & Research question {-}
The data is obtained from [this research](https://academic.oup.com/nar/article/47/11/5587/5485068?login=false). 

```{r, warning=FALSE, echo=FALSE}
library(knitr)
library(tidyverse)

metadata <- data.frame(
  `GEO Sample id` = c("GSM3393013", "GSM3393014", "GSM3393015", "GSM3393016"),
  `Cell line` = c("Fibroblast line 1 (CL1500023)",
                  "Fibroblast line 1 (CL1500023)",
                  "Fibroblast line 1 (CL1500023)",
                  "Fibroblast line 1 (CL1500023)"),
  Treatment = c("BCLXL", "BCLXL", "ONECUT1", "ONECUT1"),
  `Run id` = c("SRR7866699", "SRR7866700", "SRR7866701", "SRR7866702")
)

kable(metadata, caption = "Sample metadata table") 
```



Research question: Do ORA and GSEA lead to the same biological interpretation of ONECUT1-induced fibroblast transdifferation? 

Subquestions:
- Are GO terms found by ORA and GSEA the same? 
- Does neurogenesis/nervous system development still appear? 
- Do apoptotic processes remain enriched? 
- Do GSEA find subtler pathways that were missed by ORA? 
- To what extend do conclusions depend on the padj <0.01 and LFC > 1? 
- What is the difference in the results of GSEA GO term analysis and KEGG pathways analysis? 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

``` {r, libraries}
library(png)
library(grid)
library(gridExtra)
library(tidyverse)
library(DESeq2)
library(org.Hs.eg.db)
library(pheatmap)
library(GOstats)
library(AnnotationDbi) 
```

```{r}
count_table<-readRDS(file = "~/bookdown-demo/data/rna_seq/read_counts_OC1.rds")
count_table_matrix<-count_table$counts

metadata_OC1 <- read_csv("~/bookdown-demo/data/rna_seq/onecut_sampledata_OC1.csv")

# Convert the metadata to data frame object
metadata_OC1 <- as.data.frame(metadata_OC1)

# Add rownames to the metadata data frame
rownames(metadata_OC1) <- paste0(metadata_OC1$Run, ".bam")

# Show first lines of metadata object
# head(metadata_OC1)

# Check if column names of count table are the same as row names of metadata object
# colnames(count_table_matrix) == rownames(metadata_OC1)

# verkort treatment naam i.v.m. heatmap en andere grafieken
metadata_OC1 <- metadata_OC1 %>%
  mutate(
    treatment = str_replace_all(Cell_type, " ", "_"), # replace spaces with underscores
    treatment = case_when(                               
      treatment == "Skin_derived_fibroblast_overexpressing_Bclxl" ~ "Bclxl_control",
      treatment == "2_days_after_induction_of_OC1_in_skin_derived_fibroblasts" ~ "OC1_treatment",
      TRUE ~ treatment
    )
  )

# factor maken
metadata_OC1$treatment<- factor(
  metadata_OC1$treatment,
  levels = c("Bclxl_control","OC1_treatment")
)

# Create the DESeqDataSet object
dds <- DESeqDataSetFromMatrix(
  countData = count_table_matrix,
  colData = metadata_OC1, 
  design = ~ treatment
)

# Perform the DGE analysis using DESeq2
Trans_diff_dge <- DESeq(dds)
Trans_diff_dge_results <- results(Trans_diff_dge)

```



``` {r}
trans_diff_dge_go <- results(Trans_diff_dge, alpha = 0.01, lfcThreshold = 1) 

#upregulated genes 

upregulated_genes <- trans_diff_dge_go %>% data.frame() %>% filter(log2FoldChange > 1, padj < 0.01) %>% rownames()

#downregulated genes 
downregulated_genes <- trans_diff_dge_go %>% data.frame() %>% filter(log2FoldChange < -1, padj < 0.01) %>% rownames()

all_genes <- trans_diff_dge_go %>% data.frame() %>% rownames()

# Perform GO term enrichment analysis for upregulated genes 
test_object_upreg <- new("GOHyperGParams",
                   geneIds = upregulated_genes,
                   universeGeneIds = all_genes, 
                   annotation = "org.Hs.eg.db", 
                   ontology = "BP", 
                   pvalueCutoff = 1,
                   testDirection = "over")
goterm_analysis_upreg <- hyperGTest(test_object_upreg)

goterm_analysis_upreg_results <- summary(goterm_analysis_upreg)

```

``` {r}
# Perform GO term enrichment analysis for downregulated genes
test_object_downreg <- new("GOHyperGParams",
                   geneIds = downregulated_genes,
                   universeGeneIds = all_genes, 
                   annotation = "org.Hs.eg.db", 
                   ontology = "BP", 
                   pvalueCutoff = 1,
                   testDirection = "over")

goterm_analysis_downreg <- hyperGTest(test_object_downreg)

goterm_analysis_downreg_results <- summary(goterm_analysis_downreg)

#Adjust p values for multiple testing 
goterm_analysis_upreg_results$padj <- p.adjust(goterm_analysis_upreg_results$Pvalue, method = "BH")

goterm_analysis_downreg_results$padj <- p.adjust(goterm_analysis_downreg_results$Pvalue, method = "BH")

# Select only gene sets that are larger than 5 but smaller 
# than 500 (to prevent taking along very small and very large
# gene sets).
goterm_analysis_upreg_results <- goterm_analysis_upreg_results %>% filter(Count > 5) %>% filter(Count < 500)

goterm_analysis_downreg_results <- goterm_analysis_downreg_results %>% filter(Count > 5) %>% filter(Count < 500)

#Select the top 20 GO terms 
goterm_analysis_upreg_top20  <- goterm_analysis_upreg_results[order(goterm_analysis_upreg_results$padj)[1:20],]
goterm_analysis_downreg_top20 <- goterm_analysis_downreg_results[order(goterm_analysis_downreg_results$padj)[1:20],]


```

``` {r}

#order by padj, term is ordered using padj
goterm_analysis_upreg_top20$Term <- factor(goterm_analysis_upreg_top20$Term, 
                                     levels = goterm_analysis_upreg_top20$Term[
                                       order(goterm_analysis_upreg_top20$padj, decreasing = TRUE)])

#Plot the p-values of the top 20 GO terms (upregulated)
p1 <- goterm_analysis_upreg_top20 %>% ggplot(aes(x = Term, y =-log10(padj))) +
  geom_point() +
  coord_flip() + 
  ylab(expression(-log[10](adjusted~italic(P)~value))) +
  xlab("GO terms") + 
  ggtitle("Top 20 enriched GO terms for upregulated genes\n in ONECUT1 vs BCLXL")

p1

#order by padj, term is ordered using padj
goterm_analysis_downreg_top20$Term <- factor(goterm_analysis_downreg_top20$Term, 
                                     levels = goterm_analysis_downreg_top20$Term[
                                       order(goterm_analysis_downreg_top20$padj, decreasing = TRUE)])

#Plot the p-values of the top 20 GO terms (downregulated)
p2 <- goterm_analysis_downreg_top20 %>% ggplot(aes(x = Term, y =-log10(padj))) +
  geom_point() +
  coord_flip() + 
  ylab(expression(-log[10](adjusted~italic(P)~value))) +
  xlab("GO terms") + 
  ggtitle("Top 20 enriched GO terms for downregulated genes\n in ONECUT1 vs BCLXL")

p2

```
