# ggplot2: ravenplot {-}

A ggplot2 extension package providing fantasy-inspired themes, color palettes, and typography settings for atmospheric data visualizations.

The `ravenplot` package extends ggplot2 by providing:

- 10 color palettes: "vampire", "vampire_heatmap", "chocolate", "chocolate_heatmap", "witchy", "witchy_heatmap", "lotus", "lotus_heatmap", "tealish", "tealish_heatmap" 

``` {r, echo=FALSE, message=FALSE, warning=FALSE}

library("ravenplot")
library("tidyverse")
library(patchwork)
load_raven_fonts()

plot_palette <- function(palette_name) {

  colors <- get_raven_palette(palette_name)

  df <- data.frame(
    x = seq_along(colors),
    y = 1,
    col = colors
  )

  ggplot(df, aes(x, y, fill = col)) +
    geom_tile() +
    scale_fill_identity() +
    labs(
      title = paste("Palette:", palette_name)
    ) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5)
    )
}

p1_1 <- plot_palette("vampire")
p1_2 <- plot_palette("vampire_heatmap")

p2_1 <- plot_palette("witchy")
p2_2 <- plot_palette("witchy_heatmap")

p3_1 <- plot_palette("tealish")
p3_2 <- plot_palette("tealish_heatmap")

p4_1 <- plot_palette("lotus")
p4_2 <- plot_palette ("lotus_heatmap")

p5_1 <- plot_palette("chocolate")
p5_2 <- plot_palette("chocolate_heatmap")

(p1_1 | p1_2) / (p2_1 | p2_2) / (p3_1 | p3_2) / (p4_1 | p4_2) / (p5_1 | p5_2)

```

- 7 point styles (Unicode-based symbols): "rhombus", "sun", "heart", "circle", "square", "triangle", "diamond"

- 4 font themes: "fancy", "elegant", "simple", "standard"

- 2 background themes: "transparent", "vintage"

## Scatter plot {-}

``` {r}
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_raven_point(dot_type = "heart", size = 5) +
  scale_color_manual(values = get_raven_palette("vampire")) +
  labs(
    title = "Scatter Plot",
    subtitle = "Raven points with vampire palette",
    x = "Weight",
    y = "MPG"
  ) +
  theme_raven(
    background_type = "vintage",
    font_theme = "fancy",
    color_theme = "vampire"
  ) 
```

## Bar plot {-}

```{r}

ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  scale_fill_manual(values = get_raven_palette("tealish")) +
  labs(
    title = "Bar Plot",
    subtitle = "Vehicle classes",
    x = "Class",
    y = "Count"
  ) +
  theme_raven(
    background_type = "vintage",
    font_theme = "elegant",
    color_theme = "tealish",
    major_grid = FALSE
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Heatmap {-}

```{r}
corr <- cor(mtcars)

corr_df <- as.data.frame(as.table(corr))

ggplot(corr_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradientn(
    colours = get_raven_palette("witchy_heatmap")
  ) +
  labs(
    title = "Heatmap",
    subtitle = "Motor trend correlations",
    x = "",
    y = ""
  ) +
  theme_raven(
    background_type = "vintage",
    font_theme = "elegant",
    color_theme = "witchy",
    major_grid = FALSE
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
