# Parameterized Rmarkdown report: The number of COVID-19 cases in Europe {-}

The first part of this portfolio (“Code”) presents the code used to create the parameterized R Markdown report. The second part (“Report: The Number of COVID-19 Cases in Europe”) demonstrates the resulting parameterized report.

## Code {-}
```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE)
```

### Load libraries {-}

```{r, echo=TRUE, eval=FALSE}
library(tidyverse)
library(shiny)
library(scales)
```

### Import data from the [source](https://www.ecdc.europa.eu/en/publications-data/data-daily-new-cases-covid-19-eueea-country) {-}

```{r, echo=TRUE, eval=FALSE}
covid_data <- read.csv("https://opendata.ecdc.europa.eu/covid19/nationalcasedeath_eueea_daily_ei/csv", na.strings = "") %>%
  mutate(
    date = as.Date(dateRep, format = "%d/%m/%Y"),
    day = as.integer(day),
    month = as.integer(month),
    year = as.integer(year),
    cases = as.numeric(cases),
    deaths = as.numeric(deaths))
```

This report allows the user to choose a country, year, and one or more months.  
The summary and graph update automatically based on the selected values.

### Select input: Country, year, month(s) {-}

```{r, echo=TRUE, eval=FALSE}
selectInput(
  inputId = "country",
  label = "Choose a country:",
  choices = sort(unique(covid_data$countriesAndTerritories)),
  selected = "Netherlands"
)
selectInput(
  inputId = "year_selected",
  label = "Choose a year:",
  choices = sort(unique(covid_data$year)),
  selected = 2021
)
checkboxGroupInput(
  inputId = "months_selected",
  label = "Choose one or more months:",
  choices = setNames(1:12, month.name),
  selected = 7
)
```

### Data filtering {-}

```{r, echo=TRUE, eval=FALSE}
filtered_data <- reactive({
  req(input$country)
  req(input$year_selected)
  req(input$months_selected)

  covid_data %>%
    filter(
      countriesAndTerritories == input$country,
      year == as.numeric(input$year_selected),
      month %in% as.numeric(input$months_selected)
    ) %>%
    arrange(date)
})
```

### Summary {-}

```{r, echo=TRUE, eval=FALSE}
renderTable({
  selected_data <- filtered_data()

  tibble(
    Country = input$country,
    Year = input$year_selected,
    Months = paste(month.name[as.numeric(input$months_selected)], collapse = ", "),
    `Number of days with data` = nrow(selected_data),
    `Total COVID-19 cases` = sum(selected_data$cases, na.rm = TRUE),
    `Total COVID-19 deaths` = sum(selected_data$deaths, na.rm = TRUE)
  )
})
```

### Daily COVID-19 Cases Graph {-}

```{r, echo=TRUE, eval=FALSE}
renderPlot({
  ggplot(filtered_data(), aes(x = date, y = cases)) +
    geom_line(color = "#3EB9A8", linewidth = 1) +
    labs(
      title = paste("Daily COVID-19 Cases in", input$country),
      subtitle = paste(
        "Year:", input$year_selected,
        "| Month(s):",
        paste(month.name[as.numeric(input$months_selected)], collapse = ", ")
      ),
      x = "Date",
      y = "Number of cases"
    ) +
    theme_minimal()
})
```
### Daily COVID-19-Related Deaths Graph {-}
```{r, echo=FALSE, eval=FALSE}
renderPlot({
  ggplot(filtered_data(), aes(x = date, y = deaths)) +
    geom_line(color = "#7F1734", linewidth = 1) +
    labs(
      title = paste("Daily COVID-19 Deaths in", input$country),
      subtitle = paste(
        "Year:", input$year_selected,
        "| Month(s):",
        paste(month.name[as.numeric(input$months_selected)], collapse = ", ")
      ),
      x = "Date",
      y = "Number of deaths"
    ) +
    theme_minimal() +
    scale_x_date(date_labels = "%d %b")
})
```

### Daily COVID-19 Cases and Deaths Graph {-}
```{r, echo=FALSE, eval=FALSE}
renderPlot({

  plot_data <- filtered_data() %>%
    select(date, cases, deaths) %>%
    pivot_longer(
      cols = c(cases, deaths),
      names_to = "type",
      values_to = "count"
    ) %>%
    mutate(
      type = recode(
        type,
        cases = "COVID-19 Cases",
        deaths = "COVID-19 Deaths"
      )
    )
  ggplot(plot_data, aes(x = date, y = count, fill = type)) +
    geom_col(position = "dodge") +
    labs(
      title = paste("COVID-19 Cases and Deaths in", input$country),
      subtitle = paste(
        "Year:", input$year_selected,
        "| Month(s):",
        paste(
          month.name[as.numeric(input$months_selected)],
          collapse = ", ")
      ),
      x = "Date",
      y = "Count",
      fill = "Category"
    ) +
    scale_fill_manual(
      values = c(
        "COVID-19 Cases" = "#3EB9A8",
        "COVID-19 Deaths" = "#7F1734"
      )
    ) +
    scale_x_date(date_labels = "%d %b") +
    scale_y_continuous(labels = comma) +
    theme_minimal()
})
```

### Interpretation {-}
```{r, echo=FALSE, eval=FALSE}
renderText({
  selected_data <- filtered_data()
  paste(
    "For", input$country, "in", input$year_selected,
    "during", paste(month.name[as.numeric(input$months_selected)], collapse = ", "),
    "there are", nrow(selected_data), "days of data.",
    "The total number of confirmed COVID-19 cases is",
    comma(sum(selected_data$cases, na.rm = TRUE)),
    "and the total number of COVID-19-related deaths is",
    comma(sum(selected_data$deaths, na.rm = TRUE)), "."
  )
})
```

[Here is the report!](data/covid_report.Rmd)
